### 2017-04-20 paper/jfp-submission/semantics.tex
let's see what the JFP format looks like..

\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
$$
\forcesym : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
$$
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
%
\begin{figure*}[t]
$$
\begin{array}{lcl}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & : & v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
% \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
% \begin{cases}
%   \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
%   \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\

%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
% \end{cases} \\
\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\vhole{\thole}},
                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\vhole{\thole}} \\
  \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t}{\tsu'}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq & \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''}, \mbox{if}\ \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu}, \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'}  \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
%While a hole may map to a value that \emph{contains} another hole, \eg a
%lambda or a tree, it may not map \emph{directly} to another hole,
As the value substitution is normalized, in the first case of \forcesym\ we
do not need to \forcesym\ the result of the substitution, the sub-hole
will be narrowed when the context demands it.

@@

\begin{figure}[t]
$$
\begin{array}{lcll}
\gensym       & :   & t \times \tsu \rightarrow v \\
\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu} &  \text{if } \thole \in dom(\tsu) \\
\gen{\tint}{\tsu}   & \defeq  & n &  \text{non-det.} \\
\gen{\tbool}{\tsu}  & \defeq  & b &  \text{non-det.} \\
\gen{\tprod{t_1}{t_2}}{\tsu}  & \defeq  & \epair{\gen{t_1}{\tsu}}{\gen{t_2}{\tsu}} & \\ % \begin{subarray}{l} v_1 = \gen{t_1}{\tsu}\\ v_2 = \gen{t_2}{\tsu}\end{subarray} \\
\gen{\ttree{t}}{\tsu}  & \defeq  & tr &  \text{non-det.} \\
\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}} &  \text{\ehole, \thole are fresh} \\
\gen{\thole}{\tsu}  & \defeq & \vhole{\thole} & \text{\ehole is fresh} \\
\end{array}
$$
\caption{Generating values}
\label{fig:gen}
\end{figure}

\paragraph{Steps and Traces}
\input{operational.tex}
%
% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
% how the let rule would work for something "let rec"-y, and there's not
% function call rule). I only mention this because I can imagine a reviewer
% wondering about your ability to generate good witnesses for function
% types. This could likely be addressed in text, by a forward reference to
% Section 3.4 where higher-order functions are handled, without changing
% any of the formalisms at the last minute.
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A configuration is a triple $\triple{\estuck}{\vsu}{\tsu}$ of an
expression $e$ or the stuck term $\stuck$, a value substitution $\vsu$,
and a type substitution $\tsu$.
%
% We write \estuck to
% denote either an expression $e$ or \stuck.
%
We write $\step{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ if the state
$\triple{\estuck}{\vsu}{\tsu}$ transitions in a \emph{single step} to
$\triple{\estuck'}{\vsu'}{\tsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of configurations
$\triple{\estuck_0}{\vsu_0}{\tsu_0}, \ldots, \triple{\estuck_n}{\vsu_n}{\tsu_n}$ such that
$\forall 0 \leq i < n$, we have
$\step{\estuck_i}{\vsu_i}{\tsu_i}{\estuck_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
%
We write \steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if $\trace$ is
a trace of the form $\triple{\estuck}{\vsu}{\tsu},\ldots,\triple{\estuck'}{\vsu'}{\tsu'}$.
%
We write \steps{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if
\steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} for some trace $\trace$.

@@

\pagebreak
\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $s = \ptype{\trace}{f}$.
We show that \emph{all} types $t$ have stuck-inducing
values by splitting cases on whether $t$ is
compatible with $s$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \resolve{\vhole{\thole}}{\vsu}$ demonstrates that
  $\eapp{f}{v}$ gets stuck.
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, we can derive
  a $v$ from $\trace$ such that $\hastype{v}{t}$ and $\eapp{f}{v}$ gets
  stuck.
% \item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
%   such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
%   % \ES{do we need to say anythign else?}
\qedhere
\end{description}
\end{proof}

>>>>

\paragraph{Narrowing Types} The procedure % $\force{v}{t}{\vsu}{\tsu}$,
\[
\forcesym : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
\]
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
%
\begin{figure}[t]
\[
\begin{array}{lll}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
% \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
% \begin{cases}
%   \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
%   \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\

%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
% \end{cases} \\
\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq % \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    & \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}, \\
                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\end{array} \\
  \triple{\stuck}{\vsu}{\tsu} & \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}\end{array} \\
  \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} & \textbf{if} \begin{array}{l} \tsu' = \unify{\{\thole, t\}}{\tsu}, \\ v = \gen{t}{\tsu'} \end{array}\\
\end{cases}  \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''} & \hspace{-42mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu},\\ \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'} \end{array} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq \triple{\vleaf{t_1}}{\vsu}{\tsu'} & \hspace{-42mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array}\\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} \hspace{-3mm} & \defeq \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'} & \hspace{-42mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array} \\
\force{v}{t}{\vsu}{\tsu} & \defeq \triple{\stuck}{\vsu}{\tsu}
\end{array}
\]
\caption{Narrowing values}
\label{fig:narrow}
\end{figure}
%
%While a hole may map to a value that \emph{contains} another hole, \eg a
%lambda or a tree, it may not map \emph{directly} to another hole,
As the value substitution is normalized, in the first case of \forcesym\ we
do not need to \forcesym\ the result of the substitution, the sub-hole
will be narrowed when the context demands it.

@@

\begin{figure}[t]
\[
\begin{array}{lcll}
\gensym       & :   & t \times \tsu \rightarrow v \\
\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu} &  \text{if } \thole \in dom(\tsu) \\
\gen{\tint}{\tsu}   & \defeq  & n &  \text{non-det.} \\
\gen{\tbool}{\tsu}  & \defeq  & b &  \text{non-det.} \\
\gen{\tprod{t_1}{t_2}}{\tsu}  & \defeq  & \epair{\gen{t_1}{\tsu}}{\gen{t_2}{\tsu}} & \\ % \begin{subarray}{l} v_1 = \gen{t_1}{\tsu}\\ v_2 = \gen{t_2}{\tsu}\end{subarray} \\
\gen{\ttree{t}}{\tsu}  & \defeq  & tr &  \text{non-det.} \\
\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}} &  \text{\ehole, \thole are fresh} \\
\gen{\thole}{\tsu}  & \defeq & \vhole{\thole} & \text{\ehole is fresh} \\
\end{array}
\]
\caption{Generating values}
\label{fig:gen}
\end{figure}

\paragraph{Steps and Traces}
\input{operational.tex}
%
% WRW notes that Figure 4 does not seem to handle recursion (it's not clear
% how the let rule would work for something "let rec"-y, and there's not
% function call rule). I only mention this because I can imagine a reviewer
% wondering about your ability to generate good witnesses for function
% types. This could likely be addressed in text, by a forward reference to
% Section 3.4 where higher-order functions are handled, without changing
% any of the formalisms at the last minute.
Figure~\ref{fig:operational} describes the small-step contextual
reduction semantics for \lang.
%
A configuration is a triple $\triple{\estuck}{\vsu}{\tsu}$ of an
expression $e$ or the stuck term $\stuck$, a value substitution $\vsu$,
and a type substitution $\tsu$.
%
% We write \estuck to
% denote either an expression $e$ or \stuck.
%
We write $\step{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ if the state
$\triple{\estuck}{\vsu}{\tsu}$ transitions in a \emph{single step} to
$\triple{\estuck'}{\vsu'}{\tsu'}$.
%
A (finite) \emph{trace} $\trace$ is a sequence of configurations
$\triple{\estuck_0}{\vsu_0}{\tsu_0}, \ldots, \triple{\estuck_n}{\vsu_n}{\tsu_n}$ such that
$\forall 0 \leq i < n$, we have
$\step{\estuck_i}{\vsu_i}{\tsu_i}{\estuck_{i+1}}{\vsu_{i+1}}{\tsu_{i+1}}$.
%
We write $\steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ if $\trace$ is
a trace of the form $\triple{\estuck}{\vsu}{\tsu},\ldots,$ $\triple{\estuck'}{\vsu'}{\tsu'}$.
%
We write \steps{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} if
\steptr{\trace}{\estuck}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'} for some trace $\trace$.

@@

\begin{proof}[\textbf{Proof of Theorem~\ref{thm:soundness}}]
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $s = \ptype{\trace}{f}$.
We show that \emph{all} types $t$ have stuck-inducing
values by splitting cases on whether $t$ is
compatible with $s$. %the partial type upto $\trace$.
%
\begin{description}
\item [Case \tcompat{s}{t}:]
  Let $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{\stuck}{\vsu}{\tsu}$.
  %
  The value $v = \resolve{\vhole{\thole}}{\vsu}$ demonstrates that
  $\eapp{f}{v}$ gets stuck.
\item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, we can derive
  a $v$ from $\trace$ such that $\hastype{v}{t}$ and $\eapp{f}{v}$ gets
  stuck.
% \item [Case \tincompat{s}{t}:] By Lemma~\ref{lem:k-stuck}, every $v$
%   such that \hastype{v}{t} demonstrates that $\eapp{f}{v}$ gets stuck.
%   % \ES{do we need to say anythign else?}
% \qedhere
\end{description}
\end{proof}
